Trends from 30-Year Observations of Downward Solar Irradiance in Thessaloniki, \colorbox{green}{Greece}

### 2023-12-27 manuscript.v5/applsci-2743235.tex
Commit 2023-12-27 10:14

% Authors, for the paper (add full first names)
\Author{Athanasios
Natsis \href{https://orcid.org/0000-0002-5199-4119}
{\orcidicon}, Alkiviadis Bais * and Charikleia Meleti}

@@

There is a consensus among researchers that the major factor affecting
the variability of SDR attenuation is the interactions of solar
radiation with atmospheric aerosols and clouds. Those interactions,
among other factors, have been analyzed with models
\citep{Li2016, Samset2018}, showing the existence of feedback mechanisms
between the two. Similar findings have been shown from the analysis of
observations at other \hl{locations} %MDPI: We revised citation. Please confirm.
 \citep{Schwarz2020, Ohvril2009, Zerefos2009, Xia2007} [and references
therein]. In the
Mediterranean region, aerosols have been recognized as an important
factor affecting the penetration of solar radiation at the surface
\citep{Fountoulakis2016, Siomos2018, Gkikas2013, Lozano2021}. These
studies investigated the long-term trend in aerosol optical depth, which
has been found to decrease in the last three decades, the transport and
composition of aerosols, and their radiative effects.

@@

\begin{enumerate}
\def\labelenumi{(\alph{enumi})}
\tightlist
\item
  \hl{Mean} %MDPI: We removed italics of units. Please confirm.
 of the measured \(\overline{\text{SDR}}_i\) (Equation~\eqref{eq:MeanVIP}): \begin{equation}
  0.91 \times \overline{\text{SDR}}_{\text{CSref},i} - 20\,\text{Wm}^{-2}
  < \overline{\text{SDR}}_i <
  1.095 \times \overline{\text{SDR}}_{\text{CSref},i} + 30\,\text{Wm}^{-2}
  \label{eq:MeanVIP}
  \end{equation}
\item
  Maximum measured value \(M_{i}\) (Equation~\eqref{eq:MaxVIP}):
  \begin{equation}
  1 \times M_{\text{CSref},i} - 75\,\text{Wm}^{-2}
  < M_{i} <
  1 \times M_{\text{CSref},i} + 75\,\text{Wm}^{-2}
  \label{eq:MaxVIP}
  \end{equation}
\item
  Length \(L_i\) of the sequential line segments, connecting the points
  of the \(11\) SDR values (Equation~\eqref{eq:VILeq}): \begin{equation}
  L_i = \sum_{i=1}^{n-1}\sqrt{\left ( \text{SDR}_{i+1} - \text{SDR}_{i}\right )^2 + \left ( t_{i+1} - t_i \right )^2}
  \label{eq:VILeq}
  \end{equation} \begin{equation}
  1 \times L_{\text{CSref},i} - 5 < L_i < 1.3 \times L_{\text{CSref},i} + 13
  \label{eq:VILcr}
  \end{equation} where \(t_i\) is the time stamp of each SDR
  measurement.
\item
  Standard deviation \(\sigma_i\) of the slope (\(s_i\)) between the
  \(11\) sequential points, normalized by the mean
  \(\overline{\text{SDR}}_i\) (Equation~\eqref{eq:VCT1}): \begin{gather}
    \sigma_i = \frac {\sqrt{\frac{1}{n-1} \sum_{i=1}^{n-1} \left( s_i - \bar{s} \right)^2}} {\overline{\text{SDR}}_i} \label{eq:VCT1} \\
    s_i = \frac{\text{SDR}_{i+1} - \text{SDR}_{i}}{t_{i+1} - t_i},\;\;   \bar{s} = \frac{1}{n-1} \sum_{i=1}^{n-1} s_i,\;\;\forall i \in \left \{ 1, 2, \ldots, n-1 \right \}\;\;
  \end{gather} For this criterion, \(\sigma_i\) should be below a
  certain threshold (Equation~\eqref{eq:VCTcr}): \begin{equation}
    \sigma_i < \ensuremath{1.1\times 10^{-4}} \label{eq:VCTcr}
  \end{equation}
\item
  Maximum difference \(X_i\) between the change in measured irradiance
  and the change in clear sky irradiance over each measurement interval:
  \begin{gather}
    X_i = \max{\left \{ \left | x_i - x_{\text{CSref},i} \right | \right \}} \label{eq:VSM3} \\
    x_i = \text{SDR}_{i+1} - \text{SDR}_{i} \forall i \in \left \{ 1, 2, \ldots, n-1 \right \} \label{eq:VSM1} \\
    x_{\text{CSref},i} = \text{SDR}_{\text{CSref},i+1} - \text{SDR}_{\text{CSref},i} \forall i \in \left \{ 1, 2, \ldots, n-1 \right \} \label{eq:VSM2}
  \end{gather} For this criterion, \(X_i\) should be below a certain
  threshold (Equation~\eqref{eq:VSMcr}): \begin{equation}
    X_i < 7.5\,\text{Wm}^{-2} \label{eq:VSMcr}
  \end{equation}
\end{enumerate}

@@

We calculated the linear trends of SDR from the departures of the mean
daily values from the daily climatology and for the three sky
conditions. These are presented in Table~\ref{tab:trendtable}, which also contains the \(2\sigma\) standard error, the Pearson's correlation
coefficient R, and the trend in absolute units. In
Figure~\ref{fig:trendALL}, we present only the time series under all-sky
conditions; the plots for clear-sky and cloud-sky conditions, are shown
in \mbox{\hl{Appendix} %MDPI: We revised citation. Please confirm.
 \ref{app1}} \mbox{(Figures~\ref{fig:trendCLEAR} and~
\ref{fig:trendCLOUD}).} In the studied period, there is no significant
break or change in the variability pattern of the time series. The
linear trends in all three datasets are positive and \hl{around} %MDPI: We removed italics of units. Please confirm. The same below.
\(0.4\%/\)y for all-sky and cloudy-sky conditions, whereas for
clear-skies the trend is much smaller (\textasciitilde{}\(0.1\%/\)y).
The linear trends were calculated taking into account the
autocorrelation of the time series, and all three are statistically
significant at least at the \(95\%\) confidence level, as they are
larger than the corresponding \(2\sigma\) errors, despite the small
values of R, which is due to the large variability of the daily values.
The clear-sky trend is very small, suggesting a small effect from
aerosols and water vapor, which are the dominant factors of the SDR
variability \citep{Fountoulakis2016, Siomos2018, Yu2022}. In contrast,
the large positive trend of SDR under cloudy skies can be attributed to
reduction in cloud cover and/or cloud optical depth. Lack of continuous
observations of cloud optical thickness that could support these
findings does not allow drawing firm conclusions. However, there are
indications that the total cloud cover as inferred from the ERA5
analysis for the grid point of Thessaloniki is decreasing over the
period of study. From the difference between all-sky and clear-sky SDR
trends, expressed in W/m$^2$/y using the long-term mean of the
respective datasets, the radiative effect of clouds is estimated to
0.96 W/m$^2$/y. This estimate is similar to the cloud radiative
forcing of 1.22 W/m$^2$/y  reported for Granada, Spain
\citep{Lozano2023}.

@@

\begin{figure}[H]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
       % \subfloat[\label{fig:cusummonth-1}]
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-1}}\hfill
       % \subfloat[\label{fig:cusummonth-2}]
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-5}}\hfill
        \subfloat[\label{fig:cusummonth-3}]
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-9}}\hfill
        }
\end{adjustwidth}
\caption{\hl{Cumulative} %MDPI: We moved the subfigure explanations into the figure caption. Please confirm.
 sum plots of the monthly SDR anomalies in (\%) for different sky conditions: (\textbf{a})~all skies; (\textbf{b}) clear skies; (\textbf{c}) cloudy skies.}\label{fig:cusummonth}

@@

\end{adjustwidth}
        \caption{\hl{Cumulative} %MDPI: We moved the subfigure explanations into the figure caption. Please confirm.
 sum plots of monthly SDR anomalies in (\%) for different sky conditions after removing the long-term linear trend: (\textbf{a})~all skies; (\textbf{b}) clear skies; (\textbf{c}) cloudy skies.}\label{fig:cusumnotrendmonthly}
\end{figure}

@@

    \end{adjustwidth}
        \caption{\hl{Long} %MDPI: We removed duplicate explanation in figure.
 -term trends of daily SDR as a function of SZA for (\textbf{a}) all-sky, (\textbf{b}) clear-sky and (\textbf{c}) cloudy-sky conditions, separately for morning and afternoon periods. Solid shapes represent statistically significant trends ($p < 0.005$).}\label{fig:szatrends}
\end{figure}

@@

\begin{figure}[H]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            \includegraphics[width=1\linewidth]{./images/SeasonalMTrendsTogether3-2}   % Seasonal from Monthly
        }

    \end{adjustwidth}
 \caption{\colorbox{red}{Monthly} %MDPI: Please add the explanation for subfifgures in the figure.
 mean anomalies of SDR by season (rows of plots) for the three sky conditions (columns of plots). The black lines represent the linear trends.}\label{fig:seasonalALL}
\end{figure}

@@

\informedconsent{\hl{  } %MDPI: Any research article describing a study involving humans should contain this statement. Please add ``Informed consent was obtained from all subjects involved in the study.'' OR ``Patient consent was waived due to REASON (please provide a detailed justification).'' OR ``Not applicable'' for studies not involving humans. You might also choose to exclude this statement if the study did not involve humans.

%Written informed consent for publication must be obtained from participating patients who can be identified (including by the patients themselves). Please state ``Written informed consent has been obtained from the patient(s) to publish this paper'' if applicable.
}

@@

\conflictsofinterest{\hl{  } %MDPI: Declare conflicts of interest or state ``The authors declare no conflict of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results. Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript; or in the decision to publish the results must be declared in this section. If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript; or in the decision to publish the results''.
} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional

>>>>

% Authors, for the paper (add full first names)
\Author{Athanasios
Natsis \href{https://orcid.org/0000-0002-5199-4119}
{\orcidicon}, \colorbox{magenta}{Alkiviadis Bais} \href{https://orcid.org/0000-0003-3899-2001} %Natsis: added orcid  
{\orcidicon}* and Charikleia Meleti}

@@

There is a consensus among researchers that the major factor affecting
the variability of SDR attenuation is the interactions of solar
radiation with atmospheric aerosols and clouds. Those interactions,
among other factors, have been analyzed with models
\citep{Li2016, Samset2018}, showing the existence of feedback mechanisms
between the two. Similar findings have been shown from the analysis of
observations at other \hl{locations} %MDPI: We revised citation. Please confirm. 
%Natsis: Confirmed
 \citep{Schwarz2020, Ohvril2009, Zerefos2009, Xia2007} [and references
therein]. In the
Mediterranean region, aerosols have been recognized as an important
factor affecting the penetration of solar radiation at the surface
\citep{Fountoulakis2016, Siomos2018, Gkikas2013, Lozano2021}. These
studies investigated the long-term trend in aerosol optical depth, which
has been found to decrease in the last three decades, the transport and
composition of aerosols, and their radiative effects.

@@

\begin{enumerate}
\def\labelenumi{(\alph{enumi})}
\tightlist
\item
  \hl{Mean} %MDPI: We removed italics of units. Please confirm. 
  %Natsis: Confirmed
 of the measured \(\overline{\text{SDR}}_i\) (Equation~\eqref{eq:MeanVIP}): \begin{equation}
  0.91 \times \overline{\text{SDR}}_{\text{CSref},i} - 20\,\text{Wm}^{-2}
  < \overline{\text{SDR}}_i <
  1.095 \times \overline{\text{SDR}}_{\text{CSref},i} + 30\,\text{Wm}^{-2}
  \label{eq:MeanVIP}
  \end{equation}
\item
  Maximum measured value \(M_{i}\) (Equation~\eqref{eq:MaxVIP}):
  \begin{equation}
  1 \times M_{\text{CSref},i} - 75\,\text{Wm}^{-2}
  < M_{i} <
  1 \times M_{\text{CSref},i} + 75\,\text{Wm}^{-2}
  \label{eq:MaxVIP}
  \end{equation}
\item
  Length \(L_i\) of the sequential line segments, connecting the points
  of the \(11\) SDR values (Equation~\eqref{eq:VILeq}): \begin{equation}
  L_i = \sum_{i=1}^{n-1}\sqrt{\left ( \text{SDR}_{i+1} - \text{SDR}_{i}\right )^2 + \left ( t_{i+1} - t_i \right )^2}
  \label{eq:VILeq}
  \end{equation} \begin{equation}
  1 \times L_{\text{CSref},i} - 5 < L_i < 1.3 \times L_{\text{CSref},i} + 13
  \label{eq:VILcr}
  \end{equation} where \(t_i\) is the time stamp of each SDR
  measurement.
\item
  Standard deviation \(\sigma_i\) of the slope (\(s_i\)) between the
  \(11\) sequential points, normalized by the mean
  \(\overline{\text{SDR}}_i\) (Equation~\eqref{eq:VCT1}): \begin{gather}
    \sigma_i = \frac {\sqrt{\frac{1}{n-1} \sum_{i=1}^{n-1} \left( s_i - \bar{s} \right)^2}} {\overline{\text{SDR}}_i} \label{eq:VCT1} \\
    s_i = \frac{\text{SDR}_{i+1} - \text{SDR}_{i}}{t_{i+1} - t_i},\;\;   \bar{s} = \frac{1}{n-1} \sum_{i=1}^{n-1} s_i,\;\;\forall i \in \left \{ 1, 2, \ldots, n-1 \right \}\;\;
  \end{gather} For this criterion, \(\sigma_i\) should be below a
  certain threshold (Equation~\eqref{eq:VCTcr}): \begin{equation}
    \sigma_i < \ensuremath{1.1\times 10^{-4}} \label{eq:VCTcr}
  \end{equation}
\item
  Maximum difference \(X_i\) between the change in measured irradiance
  and the change in clear sky irradiance over each measurement interval:
  \begin{gather}
    X_i = \max{\left \{ \left | x_i - x_{\text{CSref},i} \right | \right \}} \label{eq:VSM3} \\
    x_i = \text{SDR}_{i+1} - \text{SDR}_{i} \forall i \in \left \{ 1, 2, \ldots, n-1 \right \} \label{eq:VSM1} \\
    x_{\text{CSref},i} = \text{SDR}_{\text{CSref},i+1} - \text{SDR}_{\text{CSref},i} \forall i \in \left \{ 1, 2, \ldots, n-1 \right \} \label{eq:VSM2}
  \end{gather} For this criterion, \(X_i\) should be below a certain
  threshold (Equation~\eqref{eq:VSMcr}): \begin{equation}
    X_i < 7.5\,\text{Wm}^{-2} \label{eq:VSMcr}
  \end{equation}
\end{enumerate}

@@

We calculated the linear trends of SDR from the departures of the mean
daily values from the daily climatology and for the three sky
conditions. These are presented in Table~\ref{tab:trendtable}, which also contains the \(2\sigma\) standard error, the Pearson's correlation
coefficient R, and the trend in absolute units. In
Figure~\ref{fig:trendALL}, we present only the time series under all-sky
conditions; the plots for clear-sky and cloud-sky conditions, are shown
in \mbox{\hl{Appendix} %MDPI: We revised citation. Please confirm. 
	%Natsis: confirmed
 \ref{app1}} \mbox{(Figures~\ref{fig:trendCLEAR} and~
\ref{fig:trendCLOUD}).} In the studied period, there is no significant
break or change in the variability pattern of the time series. The
linear trends in all three datasets are positive and \hl{around} %MDPI: We removed italics of units. Please confirm. The same below.  
%Natsis: confirmed
\(0.4\%/\)y for all-sky and cloudy-sky conditions, whereas for
clear-skies the trend is much smaller (\textasciitilde{}\(0.1\%/\)y).
The linear trends were calculated taking into account the
autocorrelation of the time series, and all three are statistically
significant at least at the \(95\%\) confidence level, as they are
larger than the corresponding \(2\sigma\) errors, despite the small
values of R, which is due to the large variability of the daily values.
The clear-sky trend is very small, suggesting a small effect from
aerosols and water vapor, which are the dominant factors of the SDR
variability \citep{Fountoulakis2016, Siomos2018, Yu2022}. In contrast,
the large positive trend of SDR under cloudy skies can be attributed to
reduction in cloud cover and/or cloud optical depth. Lack of continuous
observations of cloud optical thickness that could support these
findings does not allow drawing firm conclusions. However, there are
indications that the total cloud cover as inferred from the ERA5
analysis for the grid point of Thessaloniki is decreasing over the
period of study. From the difference between all-sky and clear-sky SDR
trends, expressed in W/m$^2$/y using the long-term mean of the
respective datasets, the radiative effect of clouds is estimated to
0.96 W/m$^2$/y. This estimate is similar to the cloud radiative
forcing of 1.22 W/m$^2$/y  reported for Granada, Spain
\citep{Lozano2023}.

@@

\begin{figure}[H]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
       % \subfloat[\label{fig:cusummonth-1}]
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-1}}\hfill
       % \subfloat[\label{fig:cusummonth-2}]
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-5}}\hfill
       % \subfloat[\label{fig:cusummonth-3}]  %Natsis: commented this line
            {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSum-9}}\hfill
        }
\end{adjustwidth}
\caption{\hl{Cumulative} %MDPI: We moved the subfigure explanations into the figure caption. Please confirm.  %Natsis: confirmed
 sum plots of the monthly SDR anomalies in (\%) for different sky conditions: (\textbf{a})~all skies; (\textbf{b}) clear skies; (\textbf{c}) cloudy skies.}\label{fig:cusummonth}

@@

\end{adjustwidth}
        \caption{\hl{Cumulative} %MDPI: We moved the subfigure explanations into the figure caption. Please confirm.  %Natsis: confirmed
 sum plots of monthly SDR anomalies in (\%) for different sky conditions after removing the long-term linear trend: (\textbf{a})~all skies; (\textbf{b}) clear skies; (\textbf{c}) cloudy skies.}\label{fig:cusumnotrendmonthly}
\end{figure}

@@

    \end{adjustwidth}
        \caption{\hl{Long}%MDPI: We removed duplicate explanation in figure. %Natsis: confirmed
 -term trends of daily SDR as a function of SZA for (\textbf{a}) all-sky, (\textbf{b}) clear-sky and (\textbf{c}) cloudy-sky conditions, separately for morning and afternoon periods. Solid shapes represent statistically significant trends ($p < 0.005$).}\label{fig:szatrends}
\end{figure}

@@

\begin{figure}[H]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            \includegraphics[width=1\linewidth]{./images/SeasonalMTrendsTogether3-2}   %Natsis: Image update Automn -> Autumn
        }

    \end{adjustwidth}
 \caption{Time series of \colorbox{red}{monthly} %MDPI: Please add the explanation for subfifgures in the figure.  %Natsis: Done that, and also rephrased the caption
 mean anomalies of SDR by season (rows of plots) for the three sky conditions (columns of plots). The black lines represent the linear trends, (\textbf{a}-\textbf{c}) winter, (\textbf{d}-\textbf{f}) spring, (\textbf{g}-\textbf{i}) summer, (\textbf{j}-\textbf{i}) autumn, and also (\textbf{a}, \textbf{d}, \textbf{g}, \textbf{i}) all-skies, (\textbf{b}, \textbf{e}, \textbf{h}, \textbf{k}) clear-skies and (\textbf{c}, \textbf{f}, \textbf{i}, \textbf{j}) cloudy-skies.}\label{fig:seasonalALL}
\end{figure}

@@

\informedconsent{\hl{  } %MDPI: Any research article describing a study involving humans should contain this statement. Please add ``Informed consent was obtained from all subjects involved in the study.'' OR ``Patient consent was waived due to REASON (please provide a detailed justification).'' OR ``Not applicable'' for studies not involving humans. You might also choose to exclude this statement if the study did not involve humans.
%Written informed consent for publication must be obtained from participating patients who can be identified (including by the patients themselves). Please state ``Written informed consent has been obtained from the patient(s) to publish this paper'' if applicable.
Not applicable} %Natsis: section added

@@

\conflictsofinterest{\hl{  } %MDPI: Declare conflicts of interest or state ``The authors declare no conflict of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results. Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript; or in the decision to publish the results must be declared in this section. If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript; or in the decision to publish the results''.
The authors declare no conflict of interest.} %Natsis: section added
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional
